---
title: "week_8_lab"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(palmerpenguins)
library(palmerpenguins)

#packages for cluster analysis
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```


### Intro to cluster analysis (k-means, hierarchical)

#### Part 1. K-means clustering
#### exploratory visualization
```{r}
#graph penguin data by species point color, does it look like their is an opportunity to cluster?


#bill length vs bill depth
ggplot(penguins) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7)+
  scale_color_manual(values = c("orange", "cyan4", "darkmagenta"))

```

```{r}
#flipper length vs body mass
ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species,
                 shape = sex),
             size = 3,
             alpha = 0.7)+
  scale_color_manual(values = c("orange", "cyan4", "darkmagenta"))
```


```{r}
#asking R how many clusters there should be
number_est <- NbClust(penguins[3:6], min.nc = 10, method = "kmeans")

# looking at results
number_est
```

#create a complete, scaled version of the data
We're still going to use 3 clusters and see how it does, though there may be a case here for 2 given that Adelie & chinstrap penguins are pretty similar. 
We are going to do this with *complete cases* - in other words, for the variables we're using to perform k-means clustering on penguins (bill length, bill depth, flipper length, body mass), we are *dropping any observation (row) where any of those are missing*. Keep in mind that this may not be the best option for every scenario - in other cases (e.g. when we have a large proportion of missingness), we may want to impute missing values instead.

```{r}
#drop roaws where any of the four size measurements are missing
penguins_complete <- penguins %>% 
  drop_na(bill_depth_mm, flipper_length_mm, body_mass_g, bill_length_mm)

#only keep columns with the four size measurements of interest, then scale them
penguins_scale <- penguins_complete %>% 
  select(ends_with("mm"), body_mass_g) %>% 
  scale()
```


## Running K-means
Now that we have complete, scaled data for the four size variables of interest, let's run k-means. You should know the iterative process it's running through from the Week 8 lecture.  
```{r}
penguins_km <- kmeans(penguins_scale, 3) #kmeans specificying starting with 3 groups/clusters
```


```{r}
# see what it returns (different elements returned by means function):
penguins_km$size #gives us the size of each cluster

penguins_km$cluster #what cluster each observation in penguin_scale is assigned to

#bind the cluster number to the original data used for clustering, so that we can see what clister each penguin is assigned to
penguins_cl <- data.frame(penguins_complete, cluster_no = factor(penguins_km$cluster))

#plot flipper length vs body mass, indicating which cluster each penguins is assigned to (but also showing the actual species)

ggplot(data = penguins_cl)+
  geom_point(aes(x = flipper_length_mm,
             y = body_mass_g,
             color = cluster_no,
             shape = species))
```

```{r}
ggplot(data = penguins_cl)+
  geom_point(aes(x = bill_length_mm,
             y = bill_depth_mm,
             color = cluster_no,
             shape = species))
```

###finding counts of how many of each species are in each cluster
```{r}
#Find the counts of each species assigned to each cluster, then pivot_wider() to make it a contingency table:

penguins_cl %>% 
  count(species, cluster_no) %>% 
  pivot_wider(names_from = cluster_no, values_from = n) %>% 
  rename('Cluster 1' = '1', 'Cluster 2' = '2', 'Cluster 3' = '3')
```

### Part 2. Cluster Analysis: hierarchical
In this section, you'll be performing hierarchical cluster analysis (& making dendrograms) in R. You should have already watched the Week 8 lecture to understand agglomerative versus divisive clustering, as well as differences in linkages (complete, single, average). 

We will use the `stats::hclust()` function for agglomerative hierarchical clustering, using WorldBank environmental data (simplified), wb_env.csv.


#### reading in data and wrangling
```{r}
wb_env <-read.csv("wb_env.csv")

#keeping the world's top 20 GHG emitters
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20)

#scale data
wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% 
  scale()

#adding back in the country names as row names, not a country name column
rownames(wb_scaled) <- wb_ghg_20$name
```



### finding the Euclidean distance
Use the `stats::dist()` function to find the Euclidean distance in multivariate space between the different observations (countries):
```{r}
#compute dissimilarity values (euclidean distance)
euc_distance <- dist(wb_scaled, method = "euclidean")

#view(euc_distance)

```

### Perform hierarchical clustering by complete linkage

The `stats::hclust()` function performs hierarchical clustering, given a dissimilarity matrix (our matrix of euclidean distances), using a linkage that you specify. 

Here, let's use complete linkage (recall from lecture: clusters are merged by the smallest *maximum* distance between two observations in distinct clusters).
```{r}
#hierarchical clustering with complete linkage
hc_complete <- hclust(euc_distance, method = "complete")

plot(hc_complete, cex = 0.6, hang = -1)
```



### making a single dendogram
```{r}
hc_single <- hclust(euc_distance, method = "single")

plot(hc_single, cex = 0.6, hang = -1)
```


### Making a tanglegram to compare the two dendrograms

Let's make a **tanglegram** to compare clustering by complete and single linkage! We'll use the `dendextend::tanglegram()` function to make it. 

First, we'll convert to class `dendrogram`, then combine them into a list:
```{r}
#convert class to dendrogram
dend_complete <- as.dendrogram(hc_complete)

dend_single <- as.dendrogram(hc_single)
```

##### make a tanglegram
```{r}
tanglegram(dend_complete, dend_single)
```

### plotting tanglegram with ggplot

Here's how you can make your dendrogram with `ggplot` (here, I'll use the complete linkage example stored as `hc_complete`) using `ggdendrogram()`, a `ggplot` wrapper: 

```{r}
ggdendrogram(hc_complete,
             rotate = TRUE)+
  theme_minimal()+
  labs(x = "Country")
```






